Abstract Data Type Inference

### 2015-03-25 patterns.tex
some final typos

\section{Computing Finite ADT Representations}
\label{sec:patterns}

@@

We first prove that the embedding order, defined hereafter, is a wqo on width-bounded
labeled interval orders.

Let $\Sigma$ be a finite alphabet. A labeled interval order is a triple $(A,\leq,\ell)$, where $(A,\leq)$ is an interval 
order, and $\ell:A\rightarrow \Sigma$ is a labeling function. The embedding order $\subseteq$ between labeled interval orders is defined as
usual by:
\begin{itemize}
	\item $(A_1,\leq_1,\ell_1)\subseteq (A_2,\leq_2,\ell_2)$ iff there exists an injective function $h:A_1\rightarrow A_2$ that 
	preserves the labeling, i.e., $\ell_1(x)=\ell_2(h(x))$, for every $x\in A_1$, and the order constraints, i.e.,
	for every $x,y\in A_1$, if $x \leq_1 y$, then $h(x)\leq_2 h(y)$.
\end{itemize}

@@

\begin{lemma}
$\subseteq$ is a wqo on width-bounded interval orders.
\end{lemma}
\begin{proof}
By definition, for every interval order $(A,\leq,\ell)$, there exists a mapping $I$ from elements of $A$ to intervals over $\mathbb{N}$ such that for every $x,y\in A$, if $x\leq y$, then the interval $I(x)$ ends before the interval $I(y)$ (i.e., the upper bound of $I(x)$ is strictly smaller than the lower bound of $I(y)$). %In general, there are more than one mappings $I_A$ that satisfy these constraints. In the following, we assume that the mapping $I_A$ is arbitrary but fixed.
Therefore, in the following, we assume that an interval order is a multiset $\Gamma$ of triples of the form $[i,j,a]$ where $[i,j]$ is an interval on the integer line and $a$ is a symbol in $\Sigma$.

We prove that another order on interval orders, denoted by $\subseteqq$, which is stronger than the embedding order $\subseteq$, is a wqo. The order $\subseteqq$ is defined by: $\Gamma_1\subseteqq \Gamma_2$ iff there exists an injective function $h:\Gamma_1\rightarrow \Gamma_2$ such that 
\begin{enumerate}
	\item $h$ preserves the labeling, i.e., for any triple $[i,j,a]$, $h([i,j,a])=[i',j',a]$, for some $i'$, $j'$, 
	\item $h$ preserves the ordering constraints, i.e., for every two triples $[i,j,a]$ and $[k,l,b]$ such that $j< k$, $h([i,j,a])=[i',j',a]$, $h([k,l,b])=[k',l',b]$, and $j'< k'$.
	\item for every two incomparable elements $x$ and $y$ of $\Gamma_1$, if the interval of $x$ starts before the interval of $y$, then the interval of $h(x)$ also starts before the interval of $h(y)$. Formally, for every two triples $[i,j,a]$ and $[k,l,b]$ such that $i\leq k$ (i.e., the first interval starts before the second interval), $h([i,j,a])=[i',j',a]$, $h([k,l,b])=[k',l',b]$, and $i'\leq k'$.
\end{enumerate}

Assume $\Gamma_1$, $\Gamma_2$,$\ldots$ is a \emph{bad} sequence, i.e., an infinite sequence of interval orders s.t. there exists no $i < j$ with $\Gamma_i\subseteqq \Gamma_j$. Also, assume that $\Gamma_1$ is the minimal size interval order that can start a bad sequence, $\Gamma_2$ is the minimal interval order that can continue a bad sequence starting with $\Gamma_1$, and so on.

For each $\Gamma_k$, let $Min(\Gamma_k)$ be a triple $[i,j,a]$ such that 
\begin{itemize}
	\item $i$ is the minimal lower bound of an interval in $\Gamma_k$, and 
	\item $j$ is the minimal upper bound of an interval in $\Gamma_k$ with lower bound $i$.
\end{itemize}
%In general, this triple is not unique but it’s not important which one we choose.

Also, let $Init(\Gamma_k) = (Min(\Gamma_k), P_k)$, where $P_k$ is the multiset of symbols labeling elements of $\Gamma_k$ that are incomparable to $Min(\Gamma_k)$. Note that $P_k$ is bounded since we assume width-bounded interval orders.

The infinite sequence $\Gamma_1$, $\Gamma_2$,$\ldots$ contains an infinite sequence $\Gamma_{k_1}$, $\Gamma_{k_2}$,$\ldots$ which, ignoring the integer interval, have the same value for $Init$.

For each $\Gamma_k$, let $\Lambda_k$ be the interval order obtained from $\Gamma_k$ by removing the triple $Min(\Gamma_k)$.

By the minimality assumptions, the infinite sequence $\Gamma_1$, $\Gamma_2$,$\ldots$,$\Gamma_{k_1-1}$, $\Lambda_{k_1}$, $\Lambda_{k_2}$,$\ldots$ is not bad. 
Therefore, there exists $m<n$ such that $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$.

@@

Let $h$ be the injective function witnessing $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$. We prove that the extension $h'$ of $h$ between $\Gamma_{k_m}$ and $\Gamma_{k_n}$, defined by $h'(Min(\Gamma_{k_m}))=Min(\Gamma_{k_n}$ and $h'(t)=h(t)$, for all $t\in \Lambda_{k_m}$ witnesses $\Gamma_{k_m}\subseteqq \Gamma_{k_n}$. Clearly, $h'$ preserves the labeling.

To prove that $h'$ preserves the ordering constraints, let $Min(\Gamma_{k_m})=[i,j,a]$ and $[k,l,b]\in \Gamma_{k_m}$ such that $j< k$. Also, let $Min(\Gamma_{k_n})=[i',j',a]$ and $h'([k,l,b])=h([k,l,b])=[k',l',b]$. Assume by contradiction that $k'\leq j'$. Since $h$ is injective, there exists an element $[x,y,c]\in \Gamma_{k_m}$ incomparable to $Min(\Gamma_{k_m})$ which is mapped to an element $[x',y',c]$ greater than $Min(\Gamma_{k_n})$ (by definition, $\Gamma_{k_m}$ and $\Gamma_{k_n}$ have the same number of elements incomparable to $Min(\Gamma_{k_m})$ and respectively, $Min(\Gamma_{k_n})$). 
Since $[x,y,c]$ is incomparable to $[i,j,a]$, we obtain that $x\leq j< k$. Also, by the current assumptions, $k'<j'<x'$. Therefore, there exist two elements $[x,y,c]$ and $[k,l,b]$ such that the interval $[x,y]$ starts before $[k,l]$ which are mapped by $h$ to the elements $[x',y',c]$ and $[k',l',b]$ such that the interval $[x',y']$ starts after $[k',l']$. This contradicts the fact that $h$ witnesses $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$.

The properties of $Min(\Gamma_{k_m})$ and $Min(\Gamma_{k_n})$ imply that $h'$ also satisfies the third property in the definition of $\subseteqq$.

@@

The \emph{signature} of an operation $o$ in a history $h$ is the pair $\sigma(o)=(c(o),f(o))$.
Then, the \emph{signature} of a match $\mu=\set{ o' \in O : m(o') = o}$ is the pair 
\[
\sigma(\mu)=(\sigma(o),\set{\sigma(o') : m(o')=o, o'\neq o})
\]
containing the signature of the match target and the set of the signatures of the other operations.
The \emph{signature} of a history $h$ is the tuple 
\begin{align*}
\sigma(h)&=&(\set{\sigma(\mu): \mu\mbox{ a match}}, \set{\sigma(o): o\mbox{ read-only}}, \\
&& \set{\sigma(o): o\mbox{ completed and unmatched}}, \\
&& \mset{\sigma(o): o\mbox{ pending and unmatched}}).
\end{align*}

The history signature contains sets of signatures 
(for matches, read-only and respectively, completed and unmatched operations) 
and the multiset of signatures of the pending and unmatched operations.

%Assume by contradiction that $\preceq$ has an infinite antichain.
Since the set of history signatures is bounded (because the set of methods is bounded)
and the set of pending operations is also bounded (since we consider only bounded width histories),
the sequence $h_1 h_2\ldots $ contains an infinite sequence of histories $h_1' h_2' \ldots$ that have 
the same signature.

@@

The \emph{vector} of a history $h$ is the tuple
\begin{align*}
\nu(h)&=&(\mset{\sigma(\mu): \mu\mbox{ a match of $h$}}, \mset{\sigma(o): o\mbox{ read-only}}, \\
&& \mset{\sigma(o): o\mbox{ completed and unmatched}}).
\end{align*}

Let $\leq$ be an order relation on multisets $\alpha:\Sigma\rightarrow\mathbb{N}$ of (match or operation) signatures from a finite set $\Sigma$ 
defined by $\alpha\leq\alpha'$ iff $\alpha(\sigma)\leq\alpha'(\sigma)$, for every $\sigma\in\Sigma$.
Clearly, $\leq$ is a wqo and so is the component-wise extension of $\leq$ to history vectors.
Therefore, by known results, the sequence of histories $h_1' h_2' \ldots$ contains 
an infinite sequence $h_1'',h_2'',\ldots$ such that $\nu(h_1'')\leq \nu(h_2'')\leq \ldots$.

@@

For the remainder of this section, we fix an ADT $A$, and let $B$ be the
complement of $\ker A$. When $\ker A$ has bounded width, so does $B$, and thus
$\preceq$ is a wqo on $B$. Furthermore, when $A$ normal, it is closed under
$\succeq$, and thus $B$ is closed under $\preceq$. Closure under a wqo implies
representation by a finite set. Formally, we say a set $X$ is \emph{finitely
representable} if there exists a finite set $Y$ and a relation $R \subseteq Y
\times X$ such that $X = \set{ x : \exists y\in Y.\ R(y,x) }$. In our case, we
obtain a finite set from which exactly the elements of $B$ are related by
$\preceq$.

@@

  The following four histories generate the complement of the atomic
  single-value register ADT, witnessing either an unmatched read operation:
\begin{verbatim}
  [1:X] read => 1 (RO)  #
\end{verbatim}
  a read of an uninitialized register occurring after some write:
\begin{verbatim}
  [1:1] write(1)            #
  [2:2] read => empty (RO)    #
\end{verbatim}
  a read which happens before its matching write operation:
\begin{verbatim}
  [1:2] read => 1 (RO)  #
  [2:2] write(1)          #
\end{verbatim}
  and a read matching a write which is not the most recent:
\begin{verbatim}
  [1:1] write(1)        #
  [2:2] write(2)          #
  [3:1] read => 1 (RO)      #
\end{verbatim}
  Every sequential history not admitted by the atomic register ADT embeds at
  least one of these four histories.

>>>>

%!TEX root = draft.tex
\section{Computing Finite ADT Representations}
\label{sec:patterns}

@@

We first prove that the embedding order, defined hereafter, is a wqo on width-bounded
labeled interval orders.
Let $\Sigma$ be a finite alphabet. A labeled interval order is a triple $(A,\leq,\ell)$, where $(A,\leq)$ is an interval 
order, and $\ell:A\rightarrow \Sigma$ is a labeling function. The embedding order $\subseteq$ between labeled interval orders is defined as
usual by:
\begin{itemize}
	\item $(A_1,\leq_1,\ell_1)\subseteq (A_2,\leq_2,\ell_2)$ iff there exists an injective function $h:A_1\rightarrow A_2$ that 
	preserves the labeling, i.e., $\ell_1(x)=\ell_2(h(x))$, for every $x\in A_1$, and the order constraints, i.e.,
	for every $x,y\in A_1$, if $x \leq_1 y$, then $h(x)\leq_2 h(y)$.
\end{itemize}

@@

\begin{lemma}
$\subseteq$ is a wqo on width-bounded interval orders.
\end{lemma}
\begin{proof}
By definition, for every interval order $(A,\leq,\ell)$, there exists a mapping $I$ from elements of $A$ to intervals on $\mathbb{N}$ such that for every $x,y\in A$, if $x\leq y$, then the interval $I(x)$ ends before the interval $I(y)$ (i.e., the upper bound of $I(x)$ is strictly smaller than the lower bound of $I(y)$). %In general, there are more than one mappings $I_A$ that satisfy these constraints. In the following, we assume that the mapping $I_A$ is arbitrary but fixed.
Therefore, in the following, we assume that an interval order is a multiset $\Gamma$ of triples of the form $[i,j,a]$ where $[i,j]$ is an interval on $\mathbb{N}$ and $a$ is a symbol in $\Sigma$.

We actually prove that another order on interval orders, denoted by $\subseteqq$, which is stronger than the embedding order $\subseteq$, is a wqo. The order $\subseteqq$ is defined by: $\Gamma_1\subseteqq \Gamma_2$ iff there exists an injective function $h:\Gamma_1\rightarrow \Gamma_2$ such that 
\begin{enumerate}
	\item $h$ preserves the labeling, i.e., for any triple $[i,j,a]$, $h([i,j,a])=[i',j',a]$, for some $i'$ and $j'$, 
	\item $h$ preserves the ordering constraints, i.e., for every two triples $[i,j,a]$ and $[k,l,b]$ such that $j< k$, $h([i,j,a])=[i',j',a]$, $h([k,l,b])=[k',l',b]$, and $j'< k'$.
	\item for every two incomparable elements $x$ and $y$ of $\Gamma_1$, if the interval of $x$ starts before the interval of $y$, then the interval of $h(x)$ also starts before the interval of $h(y)$. Formally, for every two triples $[i,j,a]$ and $[k,l,b]$ such that $i\leq k$ (i.e., the first interval starts before the second interval), $h([i,j,a])=[i',j',a]$, $h([k,l,b])=[k',l',b]$, and $i'\leq k'$.
\end{enumerate}
We say that $h$ \emph{witnesses} $\Gamma_1\subseteqq \Gamma_2$.

Assume $\Gamma_1$, $\Gamma_2$,$\ldots$ is a \emph{bad} sequence, i.e., an infinite sequence of interval orders s.t. there exists no $i < j$ with $\Gamma_i\subseteqq \Gamma_j$. Also, assume that $\Gamma_1$ is the minimal size interval order that can start a bad sequence, $\Gamma_2$ is the minimal order that can continue a bad sequence starting with $\Gamma_1$, and so on.

For each $\Gamma_k$, let $Min(\Gamma_k)$ be a triple $[i,j,a]$ such that 
(1) $i$ is the minimal lower bound of an interval in $\Gamma_k$, and 
(2) $j$ is the minimal upper bound of an interval in $\Gamma_k$ with lower bound $i$.
%In general, this triple is not unique but it’s not important which one we choose.
Also, let $Init(\Gamma_k) = (a_k, P_k)$, where $Min(\Gamma_k)=[i,j,a_k]$, for some $i$ and $j$, and $P_k$ is the multiset of symbols labeling elements of $\Gamma_k$ that are incomparable to $Min(\Gamma_k)$. Note that $P_k$ is bounded since we assume width-bounded interval orders.

The infinite sequence $\Gamma_1$, $\Gamma_2$,$\ldots$ contains an infinite sequence $\Gamma_{k_1}$, $\Gamma_{k_2}$,$\ldots$ which have the same value for $Init$.
For each $\Gamma_k$, let $\Lambda_k$ be the interval order obtained from $\Gamma_k$ by removing the triple $Min(\Gamma_k)$.
By the minimality assumptions, the infinite sequence $\Gamma_1$, $\Gamma_2$,$\ldots$,$\Gamma_{k_1-1}$, $\Lambda_{k_1}$, $\Lambda_{k_2}$,$\ldots$ is not bad. 
Therefore, there exists $m<n$ such that $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$.

@@

Let $h$ be the injective function witnessing $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$. We prove that the extension $h'$ of $h$ between $\Gamma_{k_m}$ and $\Gamma_{k_n}$, defined by $h'(Min(\Gamma_{k_m}))=Min(\Gamma_{k_n})$ and $h'(t)=h(t)$, for all $t\in \Lambda_{k_m}$, witnesses $\Gamma_{k_m}\subseteqq \Gamma_{k_n}$.

Clearly, $h'$ preserves the labeling. 
To prove that $h'$ preserves the ordering constraints, let $Min(\Gamma_{k_m})=[i,j,a]$ and $[k,l,b]\in \Gamma_{k_m}$ such that $j< k$. Also, let $Min(\Gamma_{k_n})=[i',j',a]$ and $h'([k,l,b])=h([k,l,b])=[k',l',b]$. Assume by contradiction that $k'\leq j'$. Since $h$ is injective, there exists an element $[e,f,c]\in \Gamma_{k_m}$ incomparable to $Min(\Gamma_{k_m})$ which is mapped to an element $[e',f',c]$ greater than $Min(\Gamma_{k_n})$ (because, by definition, $\Gamma_{k_m}$ and $\Gamma_{k_n}$ have the same number of elements incomparable to $Min(\Gamma_{k_m})$ and respectively, $Min(\Gamma_{k_n})$). 
Since $[e,f,c]$ is incomparable to $[i,j,a]$, we obtain that $e\leq j< k$. Also, by the current assumptions, $k'\leq j'<e'$. Therefore, there exist two elements $[e,f,c]$ and $[k,l,b]$ such that the interval $[e,f]$ starts before $[k,l]$ which are mapped by $h$ to the elements $[e',f',c]$ and $[k',l',b]$ such that the interval $[e',f']$ starts after $[k',l']$. This contradicts the fact that $h$ witnesses $\Lambda_{k_m}\subseteqq \Lambda_{k_n}$.

The properties of $Min(\Gamma_{k_m})$ and $Min(\Gamma_{k_n})$ imply that $h'$ satisfies also the third property in the definition of $\subseteqq$.

@@

The \emph{signature} of an operation $o$ in a history $h$ is the pair $\sigma(o)=(c(o),f(o))$.
Then, the \emph{signature} of a match $\mu=\set{ o' \in O : m(o') = o}$ is the pair 
\[
\sigma(\mu)=(\sigma(o),\set{\sigma(o') : m(o')=o, o'\neq o})
\]
containing the signature of the match target and the set of the signatures of the other operations.
The \emph{signature} of a history $h$ is the tuple 
\begin{align*}
\sigma(h)&=&(\set{\sigma(\mu): \mu\mbox{ a match}}, \set{\sigma(o): o\mbox{ read-only}}, \\
&& \set{\sigma(o): o\mbox{ unmatched}}, \\
&& \mset{\sigma(o): o\mbox{ pending and non-matched}}).
\end{align*}

The history signature contains sets of signatures 
(for matches, read-only and respectively, completed and unmatched operations) 
and the multiset of signatures of the pending and non-matched operations.

%Assume by contradiction that $\preceq$ has an infinite antichain.
Since the set of history signatures is bounded (because the set of methods 
and the set of pending operations are bounded),
the sequence $h_1 h_2\ldots $ contains an infinite sequence of histories $h_1' h_2' \ldots$ that have 
the same signature.

@@

The \emph{vector} of a history $h$ is the tuple
\begin{align*}
\nu(h)&=&(\mset{\sigma(\mu): \mu\mbox{ a match of $h$}}, \mset{\sigma(o): o\mbox{ read-only}}, \\
&& \mset{\sigma(o): o\mbox{ unmatched}}).
\end{align*}

Let $\leq$ be an order relation on multisets $\alpha:\Sigma\rightarrow\mathbb{N}$ of (match or operation) signatures from a finite set $\Sigma$ 
defined by $\alpha\leq\alpha'$ iff $\alpha(\sigma)\leq\alpha'(\sigma)$, for every $\sigma\in\Sigma$.
The relation $\leq$ is a wqo and so is the component-wise extension of $\leq$ to history vectors.
Therefore, by known results, the sequence of histories $h_1' h_2' \ldots$ contains 
an infinite sequence $h_1'',h_2'',\ldots$ such that $\nu(h_1'')\leq \nu(h_2'')\leq \ldots$.

@@

For the remainder of this section, we fix an ADT $A$, and let $B$ be the
complement of $\ker A$. When $\ker A$ has bounded width, so does $B$, and thus
$\preceq$ is a wqo on $B$. Furthermore, when $A$ normal, it is closed under
$\succeq$, and thus $B$ is closed under $\preceq$~\footnote{Actually, $B$ is closed under $\preceq$ when restricted to width-bounded histories, i.e., if $h_1 \preceq h_2$, $h_1 \in B$, and $h_2$ is width-bounded, then $h_2\in B$.}.
Closure under a wqo implies
representation by a finite set. Formally, we say a set $X$ is \emph{finitely
representable} if there exists a finite set $Y$ and a relation $R \subseteq Y
\times X$ such that $X = \set{ x : \exists y\in Y.\ R(y,x) }$. In our case, we
obtain a finite set from which exactly the elements of $B$ are related by
$\preceq$.

@@

  The following four histories generate the complement of the atomic
  single-value register ADT, witnessing either an unmatched read operation:
\begin{verbatim}
  [1:X] read => 1 (RO)  #
\end{verbatim}
  a read of an uninitialized register occurring after some write:
\begin{verbatim}
  [1:1] write(1)            #
  [2:2] read => - (RO)    #
\end{verbatim}
  a read which happens before its matching write operation:
\begin{verbatim}
  [1:2] read => 1 (RO)  #
  [2:2] write(1)          #
\end{verbatim}
  and a read matching a write which is not the most recent:
\begin{verbatim}
  [1:1] write(1)        #
  [2:2] write(2)          #
  [3:1] read => 1 (RO)      #
\end{verbatim}
  Every sequential history not admitted by the atomic register ADT embeds at
  least one of these four histories.
